Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2023-02-22 paper/paper.tex
Revise section 3

We have been developing the {\em BriefCASE} toolkit to address this need.
%
\brfcs\ is a Model-Based Systems Engineering (MBSE) environment that is integrated into the Open Source AADL Tool Environment (OSATE).
%
\footnote{AADL is the acronym for Architecture Analysis and Design Language~\cite{aadl}.}
%
It adds new capabilities for building cyber-resilient systems.
%
In this paper we describe how it facilitates specifying, testing, verifying, and synthesizing high assurance components inserted into a system to improve its cyber-resiliency.

@@

As such, this paper defines a \emph{code contract} language that is a subset of the \agr\ language but with sequential evaluation semantics as opposed to \agr's existing parallel semantics.
%
The sequential semantics are requisite to code synthesis and are defined such that \agr\ is able to still prove the contract composition of the system works.
%
This paper further defines a notion of \emph{test contract} for unit testing code contracts outside of the system composition to verify input to output behavior as well as general properties of the code contract.
%
Intuitively, the behavior of a high assurance component is defined by the designer in a code contract, the designer then writes test contracts to validate that behavior by asking \agr\ to prove if the code contract ensures the output obligations in the test contract under the test contract's input assumptions.
%
Once the designer is satisfied that the code contract is correct, the system composition, with its various contracts, is verified to conclude that the system meets its cyber-requirements.

@@

Code contracts for high assurance components can be arbitrarily complex since they can have persistent state that evolves over time.
%
Test contracts allow the designer to validate code contract behavior with unit testing.
For example, one of the test contracts for the monitor is that when it sees an automation response one step after a request, then it should not alert, and it should pass the response downstream.
%
Test contracts are discussed in detail in \secref{sec:testing}.
%
\figref{fig:test} in that section is an example test contract for the added monitor in our example.

@@

Data flows through the subcomponents of a system in dependency order, with inputs being propagated to outputs through all contracts until they stabilize \ie, cannot propagate further.
%
Therefore, subcomponent contracts, and thus the top-level model, must be acyclic.
%
Cyclic systems are made acyclic by breaking cycles with delays.
%
Once the data propagation has stabilized, the model proceeds to the next input data in the input streams.
%
It is worth noting that the semantics do not model computation or communication delay: the output of one contract is seen at the input of any downstream contract in the same step of the input data stream.

@@

\subsection{Contracts and verification conditions}

@@

\agr\ uses ASCII notation for the mathematical syntax we have defined.
%
The following table spells out the mapping.
%
\[
\begin{tabular}{l|l|l}
 Operation & Math. notation & \agr \\ \hline
 followed-by &  $e \to e$          & \verb+e -> e+  \\
 logical implication &  $e \imp e$ & \verb+e => e+ \\
 if-and-only-if &  $e \iff e$      & \verb+e <=> e+ \\
 logical and &  $e \land e$      & \verb+e and e+ \\
 logical or &  $e \lor e$      & \verb+e or e+ \\
 logical not & $\neg b$      & \verb+not b+ \\
 variable definition & $v = e$     & \verb+eq v e+ \\
 Previously  & $\konst{pre}(b)$ & \verb+pre(b)+ \\
 Historically  & $\konst{Hist}(b)$ & \verb+Historically(b)+ \\
\end{tabular}
\]

@@

\subsection{Formal contract for the example system}
\label{sec:sw-contract}

@@

\begin{figure}
  \begin{center}
    \scalebox{0.62}{\usebox{\sw}}
  \end{center}
  \caption{The SW component contract.}
  \label{fig:sw}
\end{figure}

The AGREE specification for the SW component in \secref{sec:example} is given in \figref{fig:sw}.
%
\linesref{line:sw-event-def-start}{line:sw-event-def-end} define local variables that are true when data is present on the corresponding event ports for the component.  These variables are
purely for notational convenience.
%
The contract has three assumptions:
%
\begin{itemize}
\item
\lineref{line:sw-assume-1} and \lineref{line:sw-assume-2}.
%
When input is present it is well-formed.
%
The well-formed predicates are defined elsewhere using AGREE functions.

\item
\lineref{line:sw-assume-3}.
%
If there is an incoming request, it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)}, or it arrives after the component has output a start event in response to a previous request, \texttt{\textbf{Since}(\textbf{not} req, strt)}---to be read as \emph{not
request since start}.
\end{itemize}

\noindent The contract has six guarantees:
%
\begin{itemize}
\item
\lineref{line:sw-guarantee-1}, \lineref{line:sw-guarantee-2}, and \lineref{line:sw-guarantee-3}.
%
A waypoint coincides with a update on the UAV location, a start coincides with a new waypoint, and waypoints are well-formed.

\item
\lineref{line:sw-guarantee-4}.
%
A start happens with a request or one step after a request.
%
At time zero the start must coincide with the request;
%
after that, the start must coincide with the request or else is a response to a request one step earlier.
%
The guarantee works with the assumption on \lineref{line:sw-assume-3} and thus does not have to check for two requests in a row without a start, since the assumption precludes that input behavior.
%
The guarantee also does not force the start to always happen, it only says that if it does happen, it is under the defined conditions.

\item
\lineref{line:sw-guarantee-5}.
%
The alert is signaled if the start does not arrive within the one-step bound.
%
Together with \lineref{line:sw-guarantee-4} this allows for non-alerting and alerting behavior.

\item
%
\lineref{line:sw-guarantee-6}.
%
If an alert has ever happened, it continues to happen.
\end{itemize}
%
These six guarantees define the behavior of the SW component under the three assumptions and correspond to the informal descriptions given in \secref{sec:example}.

\section{Code contracts for high assurance components}
\label{sec:code-contracts}

@@

Stream-based evaluation allows values arbitrarily \emph{deep in the past} to be accessed by use of $\konst{pre}(-)$.
%
However, the programming languages targeted by SPLAT do not directly support streams.
%
We are developing a translation to eliminate such deep temporal accesses by the introduction of intermediate variables.
%
The essential property the translation needs to satisfy is a syntactic restriction, call it \emph{Pascal-like}, that supports easy translation to standard imperative languages.
%
The guiding intuition is that a Pascal-like {\Eqs} accesses at most one step in the past;
%
also, accesses to a past value of a variable are only possible until the variable is updated.
%
In informal terms, this means that ${\Eqs}\cdot G$ in a code contract has the following properties:
%
\begin{itemize}
\item Followed-by only occurs at the top level of a variable definition, \eg, $v = e_1 \to e2$;
\item $\konst{pre}(e)$ is allowed only when $e$ is a variable; and
\item $\konst{pre}(v)$ is allowed only before the defining equation $v = e$ (and may occur in $e$).
\end{itemize}
%
For example, the Fibonacci sequence $1,1,2,3,5,8,13,\ldots$ can be defined, albeit in a manner that is not Pascal-like, by
%
{\small
\begin{lstlisting}[style=agree]
  Fib = 1 -> pre(1 -> Fib + pre Fib)
\end{lstlisting}
}
%
\noindent or by (again this is not Pascal-like)
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  Fib = if N <= 1 then 1 else pre(Fib) + pre(pre(Fib))
\end{lstlisting}
}
%
\noindent or by (this time Pascal-like):
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  F2 = 42 -> pre(F1)
  F1 = 42 -> pre(Fib)
  Fib = if N <= 1 then 1 else F1 + F2
\end{lstlisting}
}
%
Please note that the initial values of \verb+F1+ and \verb+F2+ are never accessed, so 42 could be replaced by any number in their defining equations.
%
A Pascal-like ${\Eqs} \cdot G$ is straightforward to implement in conventional programming languages, as we will see in \secref{sec:synthesis}.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.62}{\usebox{\flt}} \\
    \end{tabular}
  \end{center}
  \caption{Code contract for high-assurance filter.}
  \label{fig:filter}
\end{figure}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
    \scalebox{0.62}{\usebox{\mntr}} \\
    \end{tabular}
  \end{center}
  \caption{Code contract for high-assurance monitor.}
  \label{fig:monitor}
\end{figure}

@@

A tempting approach to exercising code contracts directly is to use \splt\ to create the target binary and then use that runtime with traditional testing.
%
Such an approach adds extra steps, including test harneses, that are not needed, as it is possible to bridge the formal model in \agr, with its proof system, to traditional validation using tests.
%
The intuition for that bridge comes from the compositional reasoning in AADL and \agr.
%
Effectively, a code contract for a high assurance component should be a valid implementation of a test scenario.
%
In other words, the test scenario, as expressed in a test contract, is the desired behavior of the system, and the code contract implements that behavior.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
    \scalebox{0.60}{\usebox{\tst}} \\
    \end{tabular}
  \end{center}
  \caption{A unit test for the monitor with its test contract.}
  \label{fig:test}
\end{figure}

@@

Synthesis maps from model and specifications to code.
%
The \splt\ tool traverses the system architecture looking for occurrences of high assurance components specified by code contracts;
%
for each such occurrence it generates a CakeML program.
%
\splt\ supports three kinds of behavioral specification from which to generate code.
%
In increasing order of expressiveness, these are: regular expressions, contiguity types, and code contracts.
%
In earlier work we have described how regular expressions and contiguity types are used as specification languages for verified filter implementations.
%
In this paper, we extend this \emph{specification oriented} approach to the Turing-complete specifications of code contracts in \agr.
%
We give a short review of that previous work before going into detail on the synthesis.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\contig}}
    \end{tabular}
  \end{center}
  \caption{Filter specification.}
  \label{fig:filter-spec}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\cml}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the filter.}
  \label{fig:filter-cakeml}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \scalebox{0.60}{\usebox{\monFn}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the monitor.}
  \label{fig:monitor-cakeml}
\end{figure}

@@

CoPilot is a domain specific language for writing runtime monitors for hard real-time distributed systems embedded in Haskell and requires working Haskell knowledge to write \cite{10.1007/s11334-013-0223-x}.
%
The language semantics are defined over streams, there is an interpreter for testing the behavior of the runtime monitors, and there are two backend compilers to generate C code.
%
The generated C code is validated by generating random CoPilot programs, generating random input for each program, and checking consensus on output between the interpreter and C programs.
%
Further assurance is accomplished with bounded model checking by showing equivalence between the code from the two compiler backends over a user defined number of steps.
%
Aside from the HOL4 embedding of the code contract semantics and not requiring any HOL4 working knowledge, the work in this paper differs in that it is targeting a fully verified path from the code contract all the way to the binaries leveraging the \ckml\ compiler specifically to provide stronger guarantees on the high assurance components.

>>>>

We have been developing the {\em BriefCASE} toolkit to address this need.
%
\brfcs\ is a Model-Based Systems Engineering (MBSE) environment that is integrated into the Open Source AADL Tool Environment (OSATE).
%
\footnote{AADL is the acronym for Architecture Analysis and Design Language~\cite{aadl}.}
%
It adds new capabilities for building cyber-resilient systems.
%
In this paper we describe how it facilitates specifying, testing, verifying, and synthesizing high-assurance components inserted into a system to improve its cyber-resiliency.

@@

As such, this paper defines a \emph{code contract} language that is a subset of the \agr\ language but with sequential evaluation semantics as opposed to \agr's existing parallel semantics.
%
The sequential semantics are requisite to code synthesis and are defined such that \agr\ is able to still prove the contract composition of the system works.
%
This paper further defines a notion of \emph{test contract} for unit testing code contracts outside of the system composition to verify input to output behavior as well as general properties of the code contract.
%
Intuitively, the behavior of a high-assurance component is defined by the designer in a code contract, the designer then writes test contracts to validate that behavior by asking \agr\ to prove if the code contract ensures the output obligations in the test contract under the test contract's input assumptions.
%
Once the designer is satisfied that the code contract is correct, the system composition, with its various contracts, is verified to conclude that the system meets its cyber-requirements.

@@

Code contracts for high-assurance components can be arbitrarily complex since they can have persistent state that evolves over time.
%
Test contracts allow the designer to validate code contract behavior with unit testing.
For example, one of the test contracts for the monitor is that when it sees an automation response one step after a request, then it should not alert, and it should pass the response downstream.
%
Test contracts are discussed in detail in \secref{sec:testing}.
%
\figref{fig:test} in that section is an example test contract for the added monitor in our example.

@@

Data flows through the subcomponents of a system in dependency order, with inputs being propagated to outputs through all contracts until they stabilize (\ie, they cannot propagate further).
%
Therefore, subcomponent contracts, and thus the top-level model, must be acyclic.
%
Cyclic systems are made acyclic by breaking cycles with delays.
%
Once the data propagation has stabilized, the model proceeds to the next input data in the input streams.
%
It is worth noting that the semantics do not model computation or communication delay: the output of one contract is seen at the input of any downstream contract in the same step of the input data stream.

@@

\subsection{Contracts and Verification Conditions}

@@

\agr\ uses ASCII notation for the mathematical syntax we have defined as shown in the following table.
%
\[
\begin{tabular}{l|l|l}
 Operation & Math & \agr \\ \hline
 followed-by &  $e \to e$          & \verb+e -> e+  \\
 logical implication &  $e \imp e$ & \verb+e => e+ \\
 if-and-only-if &  $e \iff e$      & \verb+e <=> e+ \\
 logical and &  $e \land e$      & \verb+e and e+ \\
 logical or &  $e \lor e$      & \verb+e or e+ \\
 logical not & $\neg b$      & \verb+not b+ \\
 variable definition & $v = e$     & \verb+eq v e+ \\
 previously  & $\konst{pre}(b)$ & \verb+pre(b)+ \\
 historically  & $\konst{Hist}(b)$ & \verb+Historically(b)+ \\
\end{tabular}
\]

@@

\subsection{Formal Contract for the Example System}
\label{sec:sw-contract}

@@

\begin{figure}
  \begin{center}
    \resizebox{\columnwidth}{!}{\usebox{\sw}}
  \end{center}
  \caption{The SW component contract.}
  \label{fig:sw}
\end{figure}

The AGREE specification for the SW component in \secref{sec:example} is given in \figref{fig:sw}.
%
\linesref{line:sw-event-def-start}{line:sw-event-def-end} define local variables that are true when data is present on the corresponding event ports for the component.  These variables are
purely for notational convenience.
%
The contract has three assumptions:
%
\begin{compactitem}
\item
\lineref{line:sw-assume-1} and \lineref{line:sw-assume-2}:
%
when input is present it is well-formed.
%
The well-formed predicates are defined elsewhere using AGREE functions.

\item
\lineref{line:sw-assume-3}:
%
if there is an incoming request, it is either the very first one, \texttt{\textbf{Historically}(\textbf{not} req)}, or it arrives after the component has output a start event in response to a previous request, \texttt{\textbf{Since}(\textbf{not} req, strt)}---to be read as \emph{not
request since start}.
\end{compactitem}

\noindent The contract has six guarantees:
%
\begin{compactitem}
\item
\lineref{line:sw-guarantee-1}, \lineref{line:sw-guarantee-2}, and \lineref{line:sw-guarantee-3}:
%
a waypoint coincides with a update on the UAV location, a start coincides with a new waypoint, and waypoints are well-formed.

\item
\lineref{line:sw-guarantee-4}:
%
a start happens with a request or one step after a request.
%
At time zero the start must coincide with the request;
%
after that, the start must coincide with the request or else is a response to a request one step earlier.
%
The guarantee works with the assumption on \lineref{line:sw-assume-3} and thus does not have to check for two requests in a row without a start, since the assumption precludes that input behavior.
%
The guarantee also does not force the start to always happen, it only says that if it does happen, it is under the defined conditions.

\item
\lineref{line:sw-guarantee-5}:
%
the alert is signaled if the start does not arrive within the one-step bound.
%
Together with \lineref{line:sw-guarantee-4} this allows for non-alerting and alerting behavior.

\item
%
\lineref{line:sw-guarantee-6}:
%
if an alert has ever happened, it continues to happen.
\end{compactitem}
%
These six guarantees define the behavior of the SW component under the three assumptions and correspond to the informal descriptions given in \secref{sec:example}.

\section{Code Contracts for High-assurance Components}
\label{sec:code-contracts}

@@

Stream-based evaluation allows values arbitrarily \emph{deep in the past} to be accessed by use of $\konst{pre}(-)$.
%
However, the programming languages targeted by SPLAT do not directly support streams.
%
We are developing a translation to eliminate such deep temporal accesses by the introduction of intermediate variables.
%
The essential property the translation needs to satisfy is a syntactic restriction, call it \emph{Pascal-like}, that supports easy translation to standard imperative languages.
%
The guiding intuition is that a Pascal-like {\Eqs} accesses at most one step in the past;
%
also, accesses to a past value of a variable are only possible until the variable is updated.
%
In informal terms, this means that ${\Eqs}\cdot G$ in a code contract has the following properties:
%
\begin{compactitem}
\item Followed-by only occurs at the top level of a variable definition, \eg, $v = e_1 \to e2$;
\item $\konst{pre}(e)$ is allowed only when $e$ is a variable; and
\item $\konst{pre}(v)$ is allowed only before the defining equation $v = e$ (and may occur in $e$).
\end{compactitem}
%
For example, the Fibonacci sequence $1,1,2,3,5,8,13,\ldots$ can be defined, albeit in a manner that is not Pascal-like, by
%
{\small
\begin{lstlisting}[style=agree]
  Fib = 1 -> pre(1 -> Fib + pre Fib)
\end{lstlisting}
}
%
\noindent or by (again this is not Pascal-like)
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  Fib = if N <= 1 then 1 else pre(Fib) + pre(pre(Fib))
\end{lstlisting}
}
%
\noindent or by (this time Pascal-like):
%
{\small
\begin{lstlisting}[style=agree]
  N = 0 -> 1 + pre(N)
  F2 = 42 -> pre(F1)
  F1 = 42 -> pre(Fib)
  Fib = if N <= 1 then 1 else F1 + F2
\end{lstlisting}
}
%
Please note that the initial values of \verb+F1+ and \verb+F2+ are never accessed, so 42 could be replaced by any number in their defining equations.
%
A Pascal-like ${\Eqs} \cdot G$ is straightforward to implement in conventional programming languages, as we will see in \secref{sec:synthesis}.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\flt}} \\
    \end{tabular}
  \end{center}
  \caption{Code contract for high-assurance filter.}
  \label{fig:filter}
\end{figure}

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\mntr}} \\
    \end{tabular}
  \end{center}
  \caption{Code contract for high-assurance monitor.}
  \label{fig:monitor}
\end{figure}

@@

A tempting approach to exercising code contracts directly is to use \splt\ to create the target binary and then use that runtime with traditional testing.
%
Such an approach adds extra steps, including test harnesses, that are not needed, as it is possible to bridge the formal model in \agr, with its proof system, to traditional validation using tests.
%
The intuition for that bridge comes from the compositional reasoning in AADL and \agr.
%
Effectively, a code contract for a high-assurance component should be a valid implementation of a test scenario.
%
In other words, the test scenario, as expressed in a test contract, is the desired behavior of the system, and the code contract implements that behavior.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\tst}} \\
    \end{tabular}
  \end{center}
  \caption{A unit test for the monitor with its test contract.}
  \label{fig:test}
\end{figure}

@@

Synthesis maps from model and specifications to code.
%
The \splt\ tool traverses the system architecture looking for occurrences of high-assurance components specified by code contracts;
%
for each such occurrence it generates a CakeML program.
%
\splt\ supports three kinds of behavioral specification from which to generate code.
%
In increasing order of expressiveness, these are: regular expressions, contiguity types, and code contracts.
%
In earlier work we have described how regular expressions and contiguity types are used as specification languages for verified filter implementations.
%
In this paper, we extend this \emph{specification oriented} approach to the Turing-complete specifications of code contracts in \agr.
%
We give a short review of that previous work before going into detail on the synthesis.

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\contig}}
    \end{tabular}
  \end{center}
  \caption{Filter specification.}
  \label{fig:filter-spec}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\cml}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the filter.}
  \label{fig:filter-cakeml}
\end{figure}

@@

\begin{figure}
  \begin{center}
    \begin{tabular}{c}
      \resizebox{\columnwidth}{!}{\usebox{\monFn}}
    \end{tabular}
  \end{center}
  \caption{Synthesized CakeML for the monitor.}
  \label{fig:monitor-cakeml}
\end{figure}

@@

CoPilot is a domain specific language for writing runtime monitors for hard real-time distributed systems embedded in Haskell and requires working Haskell knowledge to write \cite{10.1007/s11334-013-0223-x}.
%
The language semantics are defined over streams, there is an interpreter for testing the behavior of the runtime monitors, and there are two backend compilers to generate C code.
%
The generated C code is validated by generating random CoPilot programs, generating random input for each program, and checking consensus on output between the interpreter and C programs.
%
Further assurance is accomplished with bounded model checking by showing equivalence between the code from the two compiler backends over a user defined number of steps.
%
Aside from the HOL4 embedding of the code contract semantics and not requiring any HOL4 working knowledge, the work in this paper differs in that it is targeting a fully verified path from the code contract all the way to the binaries leveraging the \ckml\ compiler specifically to provide stronger guarantees on the high-assurance components.
